Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-10 draft.tex
add some comments, fix some spelling and language

For the selected class of decays, we can describe that problem as follows: The PDF $P$ has a fixed dependence on the dilepton
helicity angle $\theta_1$ and the azimuthal angle $\theta_3$. (See also \refapp{btokstarll} for details
on the angular distribution.) However, at the level of decay amplitudes the dimeson system can have an arbitrarily
large total angular momentum $j$; only its third component is restricted to $j_z = -1,0,+1$.
It is then \add{convenient}\del{usual} to compute the angular distribution explicitly in terms of $\theta_1$ and $\theta_3$, but leave
the angular observables dependent on the remaining helicity angle $\theta_2$.
We advocate here that is a sensible procedure to perform an expansion in terms of Legendre polynomials $p_{l}^{(|m|)}$
with respect to the remaining angle $\theta_2$. \add{Here, the angular-momentum indices $l$ and $m$ follows from the usual
rules for addition of the angular momenta $j$ and $\tilde{j}$ of the partial wave expansion of the underlying
amplitude and its complex conjugate: $j - \tilde{j} \leq l \leq j + \tilde{j}$, and $m = j_z + \tilde{j}_z$.}

For the decay at hand, we consider the partial-wave expansion for the angular observables (see \cite{Lee:1992ih} and \refapp{partial-waves})
\begin{equation}
    S_k(\vec{\nu}) \equiv \frac{I_k(\nuvec)}{4\pi \Gamma)}\,,\quad\text{with}\quad \Gamma = I_1(\nuvec) - \frac{I_2(\nuvec)}{3}\,,
\end{equation}
\fred{It not clever to use $k$ as an index and momentum variable}
which reads
\begin{equation}
    \label{eq:expansion}
    S_{k}(\vec{\nu},\cos\theta_2) \equiv \sum_{j} \frac{1}{n_{l,|m|}} S_{k,l}(\vec{\nu}) p_{l}^{(|m|)}(\cos\theta_2)\sin\theta_2\,.
\end{equation}
The normalization factor $n_{l,|m|}$ is defined in \refeq{legendre-scalar-product}. Within our
example we have (cf. also \refapp{btokstarll} and \cite{Lee:1992ih})
\begin{equation}
    |m| = \begin{cases}
        0\,, & k = 1,2,6\,,\\
        1\,, & k = 4,5,7,8\,,\\
        2\,, & k = 3,9
    \end{cases}\,.
\end{equation}

The angular observables $S_{k,l}$ -- as defined in \refeq{expansion} --
have the merit of a well defined total angular momentum, and thus are physically discriminable.
As a consequence of the orthogonality of the Legendre polynomials, any mismodelling (or rather, lack of modelling of higher
partial-wave observables) \emph{does not affect} the method of moments as discussed in the previous section. That is to say,
ignoring further (orthogonal) terms in the PDF cannot affect $\vecest{S}$ nor its variance.\\
\fred{That's not true. $\vecest{S}$ is enlarged with more angular observables. But the first components are unaffected.}\\

@@

In what follows, we propose a systematic method to unfold all effects
of $\Ekernel$ through MC simulations and the method of moments,
using that $\Ekernel$ can be expanded --- at
least formally --- in Legendre polymials.  For illustration, we
proceed with the explicit example of a uniangular\footnote{%
  The generalization of this section to multiangular PDFs is
  straighforward. It can be achieved by promoting $x$ to a vector,
  promoting the Legendre polynomials to products of independent
  polynomials or spherical harmonics, and promoting the indices
  $i,j,k,m$ to multi-indices.  } PDF with $x = \cos\theta$. Let us
define the PDF in terms of the Legendre polymials (i.e., $f_k(x)
\equiv P_k(x)$) and angular observables $\vec{S}$ as
\begin{equation}
    P\subt(\xt) \equiv P\subt(\xt | \vec{S}) = \sum_k S_k f_k(\xt),
  \end{equation}
where $k = 0, 1, \dots$ denotes an angular-momentum-like index associated with the observables.
Normalization of $P\subt$ is equivalent to choosing $S_0 = 1/2$. Requiring $P\subt(\xt) \ge 0 \,\forall \xt$
implies $|S_k| \leq 1/2$ for $k > 0$. More stringent relations between the
$S_k$ might hold, but are of no concern here. For later use, we define
\begin{equation}
  \label{eq:def-Sk}
  S^{(m)}_k \equiv
  \begin{cases}
    1/2 \delta_{k,0}, & m=0\\
    1/2 (\delta_{k,0} + \delta_{k,m}), & m>0 \,,
  \end{cases}
\end{equation}
and note that
\begin{equation}
    P^{(m)}(\xt) \equiv P(\xt | \vec{S}^{(m)})
\end{equation}
is a valid PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
\begin{equation}
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x),\qquad \int_{-1}^{+1} \tilde{f}_k(x) f_l(x) \rmdx{x} = \delta_{k,l}\,.
\end{equation}
We now define the \emph{simulated raw} moments $\vec{Q}^{(m)}$ as
\begin{equation}
    \label{eq:analytic-raw-moments}
    Q_i^{(m)} \equiv \iint \tilde{f}_i(\xd) P^{(m)}(\xt) \Ekernel \rmdx{\xt} \rmdx{\xd},
\end{equation}
which are instrumental to our recipe. Monte Carlo estimators of these moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$, where
\begin{equation}
    x_\text{d}^{(n,m)} \sim P_E^{(m)}(\xd) \equiv \frac{1}{R^{(m)}} \int P^{(m)}(\xt) \Ekernel \rmdx{\xt}\,.
\end{equation}
In words, for each $m$ it is required to generate events from
a toy physical distribution $P^{(m)}$, for which $S_0 = S_m=1/2$ and all other
observables are set to zero. Next, propagate these events through
a detector simulation. The normalization $R^{(m)}$ is chosen such
that $\int P_E^{(m)}(\xd) \rmdx{\xd} = 1$.  We emphasize that
$R^{(m)}$ can be estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} /
N_\text{t}$, where $N_\text{t}$ corresponds to the number of
simulated true events.  The estimators then read
\begin{equation}
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}} \sum_n^{N_\text{d}} \tilde{f}_i(x_\text{d}^{(n,m)})
\end{equation}

@@

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates.
For an experimental analysis, one would now turn to the determination of the distribution of $\vecest{S}$
as a function of the detector setup. This would involve the determination of both $M$ and $\vecest{Q}$ for a number of
detector configurations, and subsequent profiling or marginalization.
While such considerations of any detector simulation are beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples \fred{Before or after thinning?} (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find that the error on the mean
of each matrix element is smaller than $6\cdot 10^{-4}$. \add{This suggests, that the so-induced
systematic error can be driven below any statistical uncertainty.}

Let us conclude this section by commenting that parts of the unfolding
matrix are universal in a sense. Specifically, they can be reused in analyses with a similar underlying decay.
Therefore, computing resources spent on improving \add{the accuracy of $\est{M}$} are not wasted.

@@

\add{We emphasize that the above results have been obtained for a flat acceptance function.
We also study the behaviour of the unfolded angular observables. However, due to limitations of computing
resources, we have to limit our study to the decay $B\to K\ell^+\ell^-$, with its three angular observables
$S_1$ to $S_3$. We express the acceptance function in terms of Legendre polynomial $P_k(x)$,
\begin{equation}
    \eps(\cos\theta) = \frac{7}{15} P_0(\cos\theta) - \frac{4}{15} P_2(\cos\theta)\,.
\end{equation}
This acceptance function approximates the one used in a recent study of the angular observables in $B\to K\ell^+\ell^-$
decays \cite{Aaij:2014tfa}. We want to focus on effects of the unfolding process itself. We therefore
use the analytic expression for the unfolding matrix, which we compute from the raw moments as
defined in \refeq{analytic-raw-moments}. We simulate $4000$ toy analyses with up to $300$ simulated events each.
We find that the previous two bullet points still hold; i.e., we do not find any bias, and the distribution
is well described by a multivariate Gaussian distribution. The latter only holds as long as the number of events
per experiment exceeds $\sim 20$.
}

@@

We thank Ulrik Egede for insightful discussions and helpful comments on the
manuscript. We are grateful to Gudrun Hiller, Martin Jung and Konstantinos Petridis for helpful discussions,
as well as Claus Grupen for helpful advice in searching the literature.
The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung (BMBF).

>>>>

For the selected class of decays, we can describe that problem as follows: The PDF $P$ has a fixed dependence on the dilepton
helicity angle $\theta_1$ and the azimuthal angle $\theta_3$. (See also \refapp{btokstarll} for details
on the angular distribution.) However, at the level of decay amplitudes the dimeson system can have an arbitrarily
large total angular momentum $j$; only its third component is restricted to $j_z = -1,0,+1$.
It is then \add{convenient}\del{usual} to compute the angular distribution explicitly in terms of $\theta_1$ and $\theta_3$, but leave
the angular observables dependent on the remaining helicity angle $\theta_2$.
We advocate here that is a sensible procedure to perform an expansion in terms of Legendre polynomials $p_{l}^{(|m|)}$
with respect to the remaining angle $\theta_2$. \add{Here, the angular-momentum indices $l$ and $m$ follows from the usual
rules for addition of the angular momenta $j$ and $\tilde{j}$ of the partial wave expansion of the underlying
amplitude and its complex conjugate: $j - \tilde{j} \leq l \leq j + \tilde{j}$, and $m = j_z + \tilde{j}_z$.}
Using the observables $S_k(\vec{\nu})$, $k=1,\dots,9$, with $\vec{\nu}=(q^2, k^2)$ \fred{It not clever to use $k$ as an index and momentum variable}
and as defined in \refapp{partial-waves}, the expansion reads
\begin{equation}
    S_{k}(\vec{\nu},\cos\theta_2) \equiv \sum_{j} \frac{1}{n_{l,|m|}} S_{k,l}(\vec{\nu}) p_{l}^{(|m|)}(\cos\theta_2)\sin\theta_2\,,
\end{equation}
where the the normalization factor $n_{l,|m|}$ is defined in \refeq{legendre-scalar-product}. Within our
example we have (cf. also \refapp{btokstarll} and \cite{Lee:1992ih})
\begin{equation}
    |m| = \begin{cases}
        0\,, & k = 1,2,6\,,\\
        1\,, & k = 4,5,7,8\,,\\
        2\,, & k = 3,9
    \end{cases}\,.
\end{equation}
The angular observables $S_{k,j}$ defined as coefficients of this
expansion in Legendre polynomials have the merit of a well defined
total angular momentum, and thus are physically distinguishable.  As a
consequence of the orthogonality of the Legendre polynomials, any
mismodeling (or rather, lack of modeling of higher partial-wave
observables) \emph{does not affect} the method of moments as discussed
in the previous section. \add{That is to say}, ignoring further
(orthogonal) terms in the PDF neither affects $\vecest{S}$ nor its
variance.
\fred{That's not true. $\vecest{S}$ is enlarged with more angular observables. But the first components are unaffected.}\\

@@

In what follows, we propose a systematic method to unfold all effects
of $\Ekernel$ through MC simulations and the method of moments,
using that $\Ekernel$ can be expanded --- at
least formally --- in Legendre polymials.  For illustration, we
proceed with the explicit example of a uniangular\footnote{%
  The generalization of this section to multiangular PDFs is
  straighforward. It can be achieved by promoting $x$ to a vector,
  promoting the Legendre polynomials to products of independent
  polynomials or spherical harmonics, and promoting the indices
  $i,j,k,m$ to multi-indices.  } PDF with $x = \cos\theta$. Let us
define the PDF in terms of the Legendre polymials (i.e., $f_k(x)
\equiv P_k(x)$) and angular observables $\vec{S}$ as
\begin{equation}
    P\subt(\xt) \equiv P\subt(\xt | \vec{S}) = \sum_k S_k f_k(\xt),
  \end{equation}
where $k = 0, 1, \dots$ denotes an angular-momentum-like index associated with the observables.
Normalization of $P\subt$ is equivalent to choosing $S_0 = 1/2$. Requiring $P\subt(\xt) \ge 0 \,\forall \xt$
implies $|S_k| \leq 1/2$ for $k > 0$. More stringent relations between the
$S_k$ might hold, but are of no concern here. For later use, we define
\begin{equation}
  \label{eq:def-Sk}
  S^{(m)}_k \equiv
  \begin{cases}
    1/2 \delta_{k,0}, & m=0\\
    1/2 (\delta_{k,0} + \delta_{k,m}), & m>0 \,,
  \end{cases}
\end{equation}
and note that
\begin{equation}
    P^{(m)}(\xt) \equiv P(\xt | \vec{S}^{(m)})
\end{equation}
is a valid PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
\begin{equation}
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x),\qquad \int_{-1}^{+1} \tilde{f}_k(x) f_l(x) \rmdx{x} = \delta_{k,l}\,.
\end{equation}
We now define the \emph{simulated raw} moments $\vec{Q}^{(m)}$ as
\begin{equation}
    \label{eq:analytic-raw-moments}
    Q_i^{(m)} \equiv \iint \tilde{f}_i(\xd) P^{(m)}(\xt) \Ekernel \rmdx{\xt} \rmdx{\xd},
\end{equation}
which are instrumental to our recipe. Monte Carlo estimators of these moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$, where
\begin{equation}
    x_\text{d}^{(n,m)} \sim P_E^{(m)}(\xd) \equiv \frac{1}{R^{(m)}} \int P^{(m)}(\xt) \Ekernel \rmdx{\xt}\,.
\end{equation}
In words, for each $m$ it is required to generate events from
a toy physical distribution $P^{(m)}$ in which $S_0 = S_m=1/2$ and all other
observables are set to zero, and to propagate these events through
a detector simulation. The normalization $R^{(m)}$ is chosen such
that $\int P_E^{(m)}(\xd) \rmdx{\xd} = 1$.  We emphasize that
$R^{(m)}$ can be estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} /
N_\text{t}$, where $N_\text{t}$ corresponds to the number of
simulated true events.  The estimators then read
\begin{equation}
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}} \sum_n^{N_\text{d}} \tilde{f}_i(x_\text{d}^{(n,m)})
\end{equation}

@@

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates.
For an experimental analysis, one would now turn to the determination of the distribution of $\vecest{S}$
as a function of the detector setup. This would involve the determination of both $M$ and $\vecest{Q}$ for a number of
detector configurations, and subsequent profiling or marginalization.
While such considerations of any detector simulation are beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples \fred{Before or after thinning?} (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find that the error on the mean is
smaller than $6\cdot 10^{-4}$. Let us conclude this section by commenting that parts of the unfolding
matrix are universal in a sense. Specifically, they can be reused in analyses with a similar underlying decay.
Therefore, computing resources spent on improving its accuracy are not wasted.

@@

\add{We emphasize that the above results have been obtained for a flat acceptance function.
We also study the behaviour of the unfolded angular observables. However, due to limitations of computing
resources, we have to limit our study to the decay $B\to K\ell^+\ell^-$, with its three angular observables
$S_1$ to $S_3$. We express the acceptance function in terms of Legendre polynomial $P_k(x)$,
\begin{equation}
    \eps(\cos\theta) = \frac{7}{15} P_0(\cos\theta) - \frac{4}{15} P_2(\cos\theta)\,.
\end{equation}
This acceptance function approximates the one used in a recent study of the angular observables in $B\to K\ell^+\ell^-$
decays \cite{Aaij:2014tfa}. We want to focus on effects of the unfolding process itself. We therefore
use the analytic expression for the unfolding matrix, which we compute from the raw moments as
defined in \refeq{analytic-raw-moments}. We simulate $4000$ toy analyses with up to $300$ simulated events each.
We find that the previous two bullet points still hold; i.e., we do not find any bias, and the distribution
is well described by a multivariate Gaussian distribution.
}

@@

We would like to thank Ulrik Egede, Gudrun Hiller, Martin Jung and Konstantinos Petridis for insightful discussions,
as well as Claus Grupen for helpful advice in searching the literatur.
The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung (BMBF).
